### 2021-11-20 text/outline/draft.tex
update simpsec objects: poisson grid [1,4]; ep_tol 1e-5 for glmeiv_slow.

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that CRISPR perturbations \blue{are assigned stochastically to cells and cannot be observed directly}. As a consequence, one cannot know with certainty which cells were perturbed. Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed. Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression. The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.

@@

Our work makes two main contributions. First, we conduct a careful study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. \blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. % Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}

@@

After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its gene expressions) and the set of perturbations that it received. Finally, the scientist conducts perturbation-to-gene association analyses. Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations. For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom). Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of the gene expression change in response to the perturbation. If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene. This procedure is repeated for a large set of preselected perturbation-gene pairs. \blue{The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.}

\subsection{Analysis challenges}
\blue{High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.} Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed. Instead, perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect proxies of perturbation presence. We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b). \blue{Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b). Technical factors act as confounders in the measurement process and therefore must be accounted for in differential expression models.} Third, the gene and gRNA data are sparse, discrete counts. Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes are mapped to cells that have not received a perturbation. This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing and alignment processes \cite{Replogle2020}. The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). \blue{Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution overall is greater for perturbed cells.} Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells. \blue{The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.}

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed 381 positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
\blue{In this section we study thresholded regression from empirical and theoretical perspectives, uncovering several limitations of the method. Gasperini and Xie both imputed perturbation identities onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini used negative binomial regression, whereas Xie used nonparametric independence testing. These two strategies pose similar challenges, but we investigate Gasperini's variant of the thresholding method, as it relates most closely to GLM-EIV.}

Let $n \in \N$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
$$m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
\begin{equation}\label{thresh_glm}
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
\end{equation}
where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term. Fit a GLM to obtain estimates of the parameters.

\item[3.] Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
\blue{We include the library size $d^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.}

The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors. Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean gene expression in cells that were perturbed to the mean gene expression in cells that were not perturbed. $\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different thresholds: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).

We reasoned that thresholded regression systematically underestimated effect sizes on the positive control pairs, especially for small thresholds (an example of \textit{attenuation bias}). For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed. This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}. Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells. Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.

We hypothesized, however, that the reduction in bias conferred by selecting a large threshold comes at the cost of increasing the variance of the estimator. To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$. We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d). We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias of the estimator decreases and variance increases.

@@

\subsection{Model}
\subsubsection*{Negative binomial model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of the negative binomial GLM, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
\end{equation}
\begin{equation}\label{glmeiv_model_2}
\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_3}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
\end{equation}
\begin{equation}\label{glmeiv_model_4}
\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because gRNA molecules are transcribed in the cell in the same way that gene transcripts are \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as
\begin{equation}\label{glmeiv_model_5}
p_i \sim \textrm{Bern}(\pi),
\end{equation} where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.

@@

After fitting the model, we perform inference on the estimated parameters. The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model. This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm to fit the model. Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix). Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data. The derivation is fairly involved; simulation studies (subsequent section) confirm the correctness of the results.

\subsection{Statistical accelerations}
A downside of the the EM algorithm (Algorithm \ref{algo:em_full}) is that it requires fitting many GLMs. Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs. (These numbers are based on an exploratory application of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations. The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another. We expect these tricks to be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).

@@

The M step of the ``reduced'' GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20). We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact, and, to the best of our knowledge, new). Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the ``reduced'' GLM-EIV algorithm. Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting two GLMs (via IRLS).

@@

\subsection{Zero-inflated model}
The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes. This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}. However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero. We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}. We derive efficient methods for estimation and inference in this model. Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.

\section{Simulation studies}

>>>>

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that CRISPR perturbations are assigned stochastically to cells and cannot be observed directly. As a consequence, one cannot know with certainty which cells were perturbed. Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed. Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression. The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.

@@

Our work makes two main contributions. First, we conduct a careful study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.

@@

After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its gene expressions) and the set of perturbations that it received. Finally, the scientist conducts perturbation-to-gene association analyses. Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations. For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom). Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of the gene expression change in response to the perturbation. If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene. This procedure is repeated for a large set of preselected perturbation-gene pairs. The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.

\subsection{Analysis challenges}
High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here. Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed. Instead, perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect proxies of perturbation presence. We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b). Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact the measurement of both gene and gRNA expressions and therefore act as confounders (Figure \ref{analysis_challenges}b). Third, the gene and gRNA data are sparse, discrete counts. Consequently, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes map to cells that have not received a perturbation. This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \cite{Replogle2020}. The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution is shifted upward for perturbed cells. Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells. The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.\ in 2019. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed 381 positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
We study the thresholding method from empirical and theoretical perspectives, highlighting several limitations of the approach. Gasperini and Xie both imputed perturbation assignments onto the cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini used negative binomial regression, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural approach.

Let $n \in \N$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
$$m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
\begin{equation}\label{thresh_glm}
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
\end{equation}
where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors. Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).

We reasoned that the thresholding method systematically underestimates effect sizes on the positive control pairs, especially when the threshold is small. For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed. This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}. Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells. Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.

We hypothesized, however, that the reduction in bias obtained by selecting a large threshold comes at the cost of increasing the variance of the estimator. To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 versus threshold = 20 for the target of inference $\beta^m_1$. We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d). We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias of the estimator decreases and variance increases.

@@

\subsection{Model}
\subsubsection*{Negative binomial model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
\end{equation}
\begin{equation}\label{glmeiv_model_2}
\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_3}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
\end{equation}
\begin{equation}\label{glmeiv_model_4}
\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as
\begin{equation}\label{glmeiv_model_5}
p_i \sim \textrm{Bern}(\pi),
\end{equation} where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.

@@

After fitting the model, we perform inference on the estimated parameters. The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model. This strategy, however, is prohibitively slow, as the data are large and the EM algorithm is iterative. Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982}. Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data. The derivation is fairly involved (see appendix); simulation studies confirm the correctness of the results (see Section \ref{sec:simulation}) .

\subsection{Statistical accelerations}
A downside of the the EM algorithm (Algorithm \ref{algo:em_full}) is that it requires fitting many GLMs. Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs. (These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations. The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another. We expect these tricks to be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).

@@

The M step of the ``reduced'' GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20). We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact, and, to the best of our knowledge, new). Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the ``reduced'' GLM-EIV algorithm. Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting only two GLMs (via IRLS).

@@

\subsection{Zero-inflated model}
The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes. This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}. However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero. We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}. We additionally derive efficient methods for estimation and inference in the zero-inflated model. Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard GLM-EIV model for the remainder of the main text.

\section{Simulation studies}\label{sec:simulation}
